\Title

### 2011-04-26 masters_thesis/thesis/paper.tex
Made changes from my committee, finalizing the thesis document

\committeechair{Kevin~Seppi}
\committeemembera{Dan~Ventura}
\committeememberb{David~Embley}

@@

However, given this new parallelization model we can relax the requirement of
exactly reproducing PSO in an attempt to produce better results.  We present
several such relaxations, including keeping the best speculative position
evaluated instead of the one corresponding to the standard behavior of PSO, and
speculating several iterations ahead instead of just one.  We show that these
methods dramatically improve the performance of parallel PSO in many cases.
}
\documentkeywords{%
Parallel algorithms, optimization methods, particle swarm optimization,
peculative decomposition
}

@@

\section{Branch Statistics}
\label{sec:statistics}

@@

\section{Recovering from Pruning Too Much}
\label{sec:wrong}

@@

The question of which branches to take in this infinite tree is an intriguing
one that we can only begin to explore here.  If the branch corresponding to
stagnation has a 75\% chance of being taken, as in the Random topology with
most functions, we could speculate three iterations ahead on that branch and
still have a 42\% chance of predicting correctly.  However, intuition would say
that perhaps it is better to hope that the particle is productive instead of
stagnant, so a branch where the particle updates its personal best might be
more fruitful to try.  In our experiments we try just one of the countless
possibilities, but one that turns out to work very well.  More work is needed
to compare the different branching possibilities on various functions.

@@

\fig{rbf} shows our results for this function.  The function value reported is
the sum squared error of the model fit.  \fig{rbf-stuck} shows the percent of
runs that reached a value for sum squared error of 55,000 by each time step,
which we designate as successful.  Many Iterations took on average 126 time
steps to reach this value, while Standard Subswarms took on average 298.  The
p-value for a t-test on this statistic is less than $10^{-8}$.  With our
implementation of parallel PSO, each time step took on average 10.06 seconds;
1.83 seconds of that was function evaluation time and the rest was overhead.
With this particular implementation, then, Many Iterations takes on average 21
minutes to reach a successful value, while Standard takes 50 minutes.

@@

We have detailed how to perform speculative evaluation in PSO in several
different parallel architectures.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  While
exactly reproducing PSO sometimes uses too many extra processors to be useful,
when we allow ourselves some freedom with the algorithm we see great
improvements over previous methods.  We have shown results that conclusively
demonstrate the superiority of our techniques for several functions over the
standard practice of adding particles to the swarm when extra processors are
available.

>>>>

\committeechair{Kevin~Seppi}
\committeemembera{Dan~Ventura}
\committeememberb{David~W.~Embley}

@@

However, given this new parallelization model we can relax the requirement of
exactly reproducing PSO in an attempt to produce better results.  We present
several such relaxations, including keeping the best speculative position
evaluated instead of the one corresponding to the standard behavior of PSO, and
speculating several iterations ahead instead of just one.  We show that these
methods dramatically improve the performance of parallel PSO in many cases,
giving speed ups of up to six times compared to previous parallelization
techniques.
}
\documentkeywords{%
Parallel algorithms, optimization methods, particle swarm optimization,
speculative decomposition
}

@@

\subsection{Branch Statistics}
\label{sec:statistics}

@@

\subsection{Recovering from Pruning Too Much}
\label{sec:wrong}

@@

The question of which branches to take in this infinite tree is an intriguing
one that we can only begin to explore here.  If the branch corresponding to
stagnation, \casepn, has a 75\% chance of being taken, as in the Random
topology with most functions, we could speculate three iterations ahead on that
branch and still have a 42\% chance of predicting correctly.  However,
intuition would say that perhaps it is better to hope that the particle is
productive instead of stagnant, so a branch where the particle updates its
personal best might be more fruitful to try.  In our experiments we try just
one of the countless possibilities, but one that turns out to work very well.
More work is needed to compare the different branching possibilities on various
functions.

@@

\fig{rbf} shows our results for this function.  The function value reported is
the sum squared error of the model fit.  \fig{rbf-stuck} shows the percent of
runs that reached a value for sum squared error of 55,000 by each time step,
which we designate as successful.  Many Iterations took on average 126 time
steps to reach this value, while Standard Subswarms took on average 298.  The
p-value for a t-test on this statistic is less than $10^{-8}$.  With our
implementation of parallel PSO, each time step took on average 10.06 seconds;
1.83 seconds of that was function evaluation time and the rest was overhead.
With this particular implementation, then, Many Iterations takes on average 21
minutes to reach a successful value, while Standard Subswarms takes 50 minutes.

@@

We have detailed how to perform speculative evaluation in PSO in several
different parallel architectures.  Using this methodology, the behavior of the
original PSO algorithm can either be exactly reproduced, two iterations at a
time, or the behavior can be modified in order to improve performance.  While
exactly reproducing PSO sometimes uses too many extra processors to be useful,
when we allow ourselves some freedom with the algorithm we see great
improvements over previous methods.  We have shown results that conclusively
demonstrate the superiority of our techniques for several functions over the
standard practice of adding particles to the swarm when extra processors are
available, giving speed ups of up to six compared to previous parallelization
techniques.
